A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-25 paper.tex
Add args and trees section

\begin{figure}
\centering
% TODO update the source of this figure to make less tall. It's taking
% up far too much vspace at in native coords at the moment.
\includegraphics[width=0.75\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example graph
to a gARG. (A) The original eARG; square nodes represent events, with
each recombination (red) showing the corresponding breakpoint.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the samples, $a$, $b$, and $c$. Dashed lines show edges that are
no longer present (in practice, nodes $g$ and $q$ would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig~\ref{fig-arg-in-pedigree}A.
Line thickness is proportional to the genomic span of each edge.
}
\end{figure}

@@

\section*{Simplification}\label{ARG_simplification}
% I think a sentence is needed here on why "so much" text is devoted to simplification.
% How does simplification connect to eARG vs gARG and practical benefits of gARG over eARG
% and why do you now talk about unary nodes and a bit polytomies? Hold my hand here please;)
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy,
retaining the ``meaningful'' topology within local trees
and maintaining the identity of nodes across different trees.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a tree node that has exactly
% should presence above be absence?
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-in-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the first tree (covering positions 30 to 40)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through \textsf{e}, \textsf{j}, \textsf{m}, \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples coalesce in a
 \emph{common} ancestor. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent local trees on the right. Since multiple parents are indicative of
recombination(s), nodes are coloured by the number of parents. and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
children.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
edge annotations.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \textsf{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.
% something which is impossible to encode in an eARG.

@@

% Needs another pass
An important consequence of simplifying local trees to remove
unary nodes is that we may lose the ability to identify
the ``subtree prune-and-regraft'' (SPR) operations
by which they differ~\citep{hein1990reconstructing,song2003on,song2006properties}.
% The effects of recombination as we move from left-to-right along
% the genome in the complete simulated gARG
% of Fig~\ref{fig-simplification}A are easy to interpret,
% but as more redundancy is removed from the graph
% the
As explained above, we can have multiple recombination and
common ancestor events stacking up, and therefore
many possible SPR operations. Deducing the SPRs between
trees is NP-hard in
general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but effective heuristics have been proposed~\citep{rasmussen2022espalier}.

>>>>

\begin{figure}
\centering
% TODO update the source of this figure to make less tall. It's taking
% up far too much vspace at in native coords at the moment.
\includegraphics[width=0.75\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example graph
to a gARG. (A) The original eARG; square nodes represent events, with
each recombination (red) showing the corresponding breakpoint.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the samples, $a$, $b$, and $c$. Dashed lines show edges that are
no longer present (in practice, nodes $g$ and $q$ would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig~\ref{fig-arg-in-pedigree}A.
Line thickness is proportional to the genomic span of each edge.
Nodes representing recombination events have been retained
for clarity, but could be removed by simplification if
desired.
}
\end{figure}

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy.
Many of the types of redundancy revolve around the presence
of ``unary'' nodes in the local trees
(a tree node that has exactly one child).

Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
backwards-in-time Wright-Fisher
model, shown both as a graph on the left, and the corresponding sequence of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the first tree (covering positions 30 to 40)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through \textsf{e}, \textsf{j}, \textsf{m}, \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples coalesce in a
 \emph{common} ancestor. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
children.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

The first level of simplification that we can perform is to remove
graph topology that is invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or these ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \textsf{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.
% something which is impossible to encode in an eARG.

@@

An important consequence of simplifying local trees to remove
unary nodes is that we lose information about recombination
events. This is related to the amount of \emph{precision} about
recombination events that we store, which is the topic of the next
section.

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is not as simple as it may seem.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.

Firstly, the internal nodes in the local trees must be labelled
(i.e.\ marked with the identity of the corresponding ARG node).
If the nodes are not labelled, we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.
% JK: Does this need to be said?
% (One may use the time of a node as a proxy for identity,
% but it is not reliable as we can have multiple nodes at the
% same time, e.g. with time discretisation.)

The precision of the information about recombination that
can be represented in the trees depends the ARG itself.
Secondly, the local trees must contain ``unary'' nodes.
[Some discussion].
With these two conditions met, there is a direct correspondance
between a gARG and the trees.

These properties have interesting consequences when we
consider the inference strategy of esimating local trees
independenty and subsequently stitching them together
into an ARG. Two recent methods take this approach,
\relate\ and
\texttt{espalier}~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local
trees into an ARG. [Relate does something fairly lightweight,
but by default imprecise.]
Espalier, on the other hand, is specifically in interested
in the recombination events and therefore attempts to infer
the precise ``subtree prune-and-regraft'' (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
using heuristics~\citep{rasmussen2022espalier}.
Inferring the SPRs between
leaf-labelled trees is NP-hard in
general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
